---
title: "passnyc"
output: html_notebook
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = T, warning = F, message = F,
  fig.width = 6,fig.asp = 0.618,out.width = "80%",fig.align = "center"
)
```

```{r load_libs}
library(tidyverse);library(plotly);library(leaflet);library(htmltools);library(magrittr)
```

```{r get_data}
schools <- 
  read_csv(
    file = "data/2016 School Explorer.csv",
    na = c("", "NA","N/A")
  )

tests <- read_csv(file = "data/D5 SHSAT Registrations and Testers.csv")

# Clean names
names(schools) <- str_replace_all(names(schools), "(|)|\\?|\\-", "")
names(schools) <- tolower(str_replace_all(names(schools), " ", "_"))
names(tests) <- tolower(str_replace_all(names(tests), " ", "_"))
```

```{r get_api_data}

library("RSocrata")

nyc_attend_12_17 <- read.socrata(
  url = "https://data.cityofnewyork.us/resource/s9bd-hden.json",
  app_token = Sys.getenv("socrata_token"),
  email     = Sys.getenv("socrata_email"),
  password  = Sys.getenv("socrata_pw")
)

nyc_attend_ytd <- read.socrata(
  url = "https://data.cityofnewyork.us/resource/5vg6-6vfs.json",
  app_token = Sys.getenv("socrata_token"),
  email     = Sys.getenv("socrata_email"),
  password  = Sys.getenv("socrata_pw")
)

nyc_sat_2012 <- read.socrata(
  url = "https://data.cityofnewyork.us/resource/734v-jeq5.json",
  app_token = Sys.getenv("socrata_token"),
  email     = Sys.getenv("socrata_email"),
  password  = Sys.getenv("socrata_pw")
)

nyc_graduate <- read.socrata(
  url = "https://data.cityofnewyork.us/resource/ns8x-c6af.json",
  app_token = Sys.getenv("socrata_token"),
  email     = Sys.getenv("socrata_email"),
  password  = Sys.getenv("socrata_pw")
)

```

```{r combine_data}

nyc_attend_ytd %<>% 
  mutate(schoolyear = "2017 - 2018") %>%
  select(absent,calmonth,gradelevel,gradesort,monthcode,present,released,rostercount,school,schoolyear)

nyc_attend <-
  nyc_attend_12_17 %>%
  bind_rows(nyc_attend_ytd) %>%
  mutate_at(
    .vars = vars(absent,gradesort,present,released,rostercount),
    .funs = funs(as.numeric)
  ) %>%
  mutate(tot = present + absent) %>%
  select(school,schoolyear,calmonth,monthcode,gradelevel,gradesort,present,absent,released,rostercount,tot) %>%
  group_by(school,schoolyear,gradelevel) %>% 
  summarize(
    absent = sum(absent),
    present = sum(present),
    tot = sum(tot),
    n_months = n_distinct(calmonth)
  ) %>%
  mutate(
    gradelevel = recode(
      gradelevel,
      `6` = "06",
      `7` = "07",
      `8` = "08",
      `9` = "09"
    )
  ) %>%
  filter(
    gradelevel %in% c("06","07","08")
    & schoolyear %in% c("2015 - 2016","2016 - 2017","2017 - 2018")
  ) %>%
  group_by(school) %>%
  summarize(
    absent = sum(absent),
    present = sum(present),
    tot = sum(tot),
    n_months = sum(n_months)
  ) %>%
  mutate(
    avg_absent = round(absent / n_months, digits = 1),
    avg_present = round(present / n_months, digits = 1),
    avg_tot = round(tot / n_months, digits = 1),
    attend_rate = round(present / tot * 100, digits = 1)
  ) %>%
  select(location_code = school, avg_tot, attend_rate)

rm(nyc_attend_12_17); rm(nyc_attend_ytd)

```


```{r clean_data}
schools <-
  schools %>%
  mutate(
    school_income_estimate = as.numeric(
      str_replace_all(school_income_estimate,"\\$|,","")
    ),
    sed_code = as.character(sed_code),
    economic_need_index = as.numeric(economic_need_index)
  ) %>%
  mutate_at(
    .vars = vars(matches("percent|%|rate")),
    .funs = funs(str_replace_all),
    pattern = "%", replacement = ""
  ) %>%
  mutate_at(
    .vars = vars(matches("percent|%|rate")),
    .funs = funs(as.numeric)
  ) %>%
  mutate(
    percent_nonwhite = 100 - percent_white
  )
  
  
```

The [Princeton Review](https://www.princetonreview.com/k12/shsat-information) notes that they "do not recommend waiting until 9th grade to take the test as these high schools have a very limited number of 10th grade seats available."

```{r subset_join}
subset_schools <-
  schools %>%
  filter(as.numeric(grade_high) %in% 6:8) %>%
  select(
    location_code, school_name, latitude, longitude, zip, grade_high, `community_school?`, economic_need_index, 
    student_attendance_rate, percent_of_students_chronically_absent,school_income_estimate, `rigorous_instruction_%`:student_achievement_rating, percent_ell, percent_nonwhite,
    starts_with("grade_6"),starts_with("grade_7"),starts_with("grade_8")
  ) 

tst <- 
  subset_schools %>% 
  left_join(nyc_attend, by = "location_code")

```


# What is the problem?

Per [NY Times article](https://www.nytimes.com/2018/06/21/nyregion/what-is-the-shsat-exam-and-why-does-it-matter.html), "Beginning in the fall of 2019, the city would set aside 20 percent of seats for low-income students who score just below the lowest cutoff score."

# Who are we trying to help?

We want to identify schools with kids who are: 

- from backgrounds that would diversify the specialty schools (e.g. immigrants, kids with disabilities, poor students, hardships)
- eligible to take the SHSAT (*meaning that the school should meet the above criteria for the upper grades prior to high school, since these are the grades where kids take the SHSAT*)
- in schools that do not have significant resources but are performing relatively well anyway
- in schools that have a positive culture
- "stand to gain the most" from services like after school programs, test preparation, mentoring, or resources for parents
- low income, according to the 2019 admission policy

## Identifying relevant variables

### Related to student demographic characteristics:

- Immigrants: `percent_ell`, `ends_with("_limited_english_proficient")`
- Economically disadvantaged: `economic_need_index`, `ends_with("_economically_disadvantaged")`. We also use the ``, which "reflects the socioeconomics of the school population."^[According to [the NYC Educator Guide](http://schools.nyc.gov/NR/rdonlyres/7B6EEB8B-D0E8-432B-9BF6-3E374958EA70/0/EducatorGuide_EMS_20131118.pdf)  It is calculated using the following formula: Economic Need Index = (Percent Temporary Housing) + (Percent 
HRA-eligible * 0.5) + (Percent Free Lunch Eligible * 0.5)]
- Disabilities: no variables were found in the schools data set, though we may be able to find other publically available data. 
- Absentees: `student_attendance_rate`, `percent_of_students_chronically_absent`
- Race/Ethnicity: While the PASSNYC problem statement indicated that more 'homogenous student body demographics' were a part of what it sought to address in developing specialized school enrollment for "a more diverse group of students", they did not explicitly identify which groups were over/under-represented.  We will keep these variables in to evaluate our model, but not use these variables to ilter schools out of the results.  Variables which identify race/ethnicity include: `percent_nonwhite`, `ends_with("_american_indian_or_alaska_native")`, `ends_with("_black_or_african_american")`, `ends_with("_hispanic_or_latino")`, `ends_with("_asian_or_pacific_islander")`, `ends_with("_white")`, `ends_with("_multiracial")`



We might want to look at schools which have a high percent of English language learners (ELL), which we can summarize by looking at the histogram below:

```{r}
subset_schools %>% plot_ly(x = ~percent_ell) %>% add_histogram()
```

However, the likelihood is that some schools with a high percent of ELL won't necessarily have a high percentage of African-American or Latino students.  So, we want to start by looking at all of the relevant variables in one view, called a heatmap:


```{r}
subset_schools %>% 
  set_rownames(.$sed_code) %>%
  select(economic_need_index:percent_nonwhite) %>%
  select(-ends_with("_rating")) %>%
  scale() %>%
  d3heatmap::d3heatmap(
    colors = viridisLite::viridis(30, direction = -1),
    k_row = 5,
    na.rm = T,
    show_grid = F
  )

```



# How are we trying to help them?

- Motivating them to study for the SHSAT
- Tutoring and test preparation for the SHSAT
- Application to specialty high schools

# Where is the best place to go to help them?

- Where there are the greatest number of kids fitting this criteria
- With students in the most advanced grades who are candidates for specialty schools (*How long does it take to establish relationships with schools?  If awhile, may want to pick schools that meet criteria for multiple upper grades in a row*)
- Can kids come from other schools if they're nearby?



```{r}
colnames(schools) %>% as.data.frame() %>% View()
```

## Find Relevant Variables




Find schools that have:

- high `percent_ell` 
- high `economic_need_index`
- high `student_attendance_rate` and low `percent_of_students_chronically_absent`
- up to 8th grade


```{r}
tst <-
subset_schools %>% 
  filter(
    grade_high == "08" 
    & economic_need_index >= 0.675
    | percent_ell >= 12.5
  ) %>%
  filter(
    percent_of_students_chronically_absent <= 20 
    & student_attendance_rate >= 80
  )

```


```{r}

subset_schools %>%
  plot_ly(
    x = ~percent_of_students_chronically_absent
  ) %>%
  add_histogram()

```


[community schools](https://www1.nyc.gov/site/communityschools/index.page)


```{r}
schools %>% 
  plot_ly(
    x=~fct_reorder(school_name,percent_ell,.desc = T),
    y=~percent_ell
  )

```

```{r}
schools %>% 
  plot_ly(
    x=~percent_ell,
    y=~economic_need_index
  ) %>%
  add_markers()
```

What is the criteria for specialty schools?
Are specialty schools merit based or lottery based schools?

# Maps/Location

```{r map}

schools %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~longitude, 
    ~latitude,
    popup = ~paste0(
      "<b>School Name:</b> ",htmlEscape(school_name),"<br/>",
      "<b>Address:</b> ",htmlEscape(`address_full`)
    ),
    stroke = FALSE,
    radius = 5,
    fillOpacity = 0.6
  )

```

# Appendix 1: Open NYC Datasets

The following datasets were considered as candidates for inclusion in the analysis:

- [Universal Pre-K School Locations](https://www.kaggle.com/new-york-city/nyc-universal-pre-k-upk-school-locations)
- [School District Breakdowns](https://www.kaggle.com/new-york-city/nyc-school-district-breakdowns)
- [SAT Results](https://www.kaggle.com/new-york-city/new-york-city-sat-results)
- [Graduation Outcomes](https://www.kaggle.com/new-york-city/new-york-city-graduation-outcomes)
- [School Attendance and Enrollment](https://www.kaggle.com/new-york-city/ny-school-attendance-and-enrollment)
- [NYPL Branch Services](https://www.kaggle.com/new-york-city/nyc-nypl-branch-services)
